The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2018-05-20 script/de/chapters/ellipticcurves.tex
CT1: New menutree pdfs: Updated CTB and copied CTB pdfs in setup\template-xx; corrected according online help files hlp_xx/menu_structure.html
CT1: Updated CT book (d+e): small typos

% $Id$
% ..............................................................................
%                  E L L I P T I S C H E  K U R V E N
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

Bei der digitalen Signatur muss man differenzieren: für die {\em
  Erstellung} einer digitalen Signatur benötigen auf Elliptischen
Kurven basierende Verfahren im Vergleich zu RSA nur gut ein Zehntel des
Rechenaufwandes (66 zu 515 Ganzzahlmultiplikationen). Siehe hierzu
Abbildung~\ref{ThousandBitMultiplications} (Quelle: J.  Merkle,
Elliptic Curve Cryptography Workshop, 2001). Betrachtet man die für eine
{\em Verifikation} durchzuführenden Rechenschritte, dreht sich dieses Bild
jedoch zu Gunsten von RSA um (112 zu 17 Ganzzahlmultiplikationen). Der 
Grund liegt darin, dass es bei Verwendung des RSA möglich ist, einen sehr 
kurzen Exponent für den öffentlichen Schlüssel zu wählen, solange der 
private Exponent nur hinreichend lang ist.

@@

Es gibt Faktorisierungsalgorithmen%
\footnote{John M. Pollard \index{Pollard, John M.} war an der 
Entwicklung vieler verschiedener Faktorisierungsalgorithmen beteiligt; auch
beim Faktorisieren mit ECC war er einer der führenden Köpfe. Als Mitarbeiter
von British Telekom hat er leider nie viel selbst publiziert. Auf der RSA
Konferenz 1999 wurde er für seine \glqq outstanding contributions in 
mathematics\grqq ~ausgezeichnet.\\
Im Jahr 1987 stellte H.W. Lenstra\index{Lenstra 1987} einen häufig genutzten
Faktorisierungsalgorithmus vor, der auf Elliptischen Kurven basiert 
(siehe \cite{Lenstra1987}).
  % Leerzeile vor schließender Footnote-Klammer zu #21 nötig, um einheitlich eine
  % Leerzeile am Ende der Fußnoten zu haben -- solange keine andere Lösung
  % für das Einrückproblem bei Fußnote #23 (TODO_LaTeX)

>>>>

% $Id$
% !Mode:: "TeX:DE"    % Setting document mode and submode for WinEdt
% ..............................................................................
%                  E L L I P T I S C H E  K U R V E N
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

Bei der digitalen Signatur muss man differenzieren: für die {\em
  Erstellung} einer digitalen Signatur benötigen auf Elliptischen
Kurven basierende Verfahren im Vergleich zu RSA nur gut ein Zehntel des
Rechenaufwands (66 zu 515 Ganzzahlmultiplikationen). Siehe hierzu
Abbildung~\ref{ThousandBitMultiplications} (Quelle: J.  Merkle,
Elliptic Curve Cryptography Workshop, 2001). Betrachtet man die für eine
{\em Verifikation} durchzuführenden Rechenschritte, dreht sich dieses Bild
jedoch zu Gunsten von RSA um (112 zu 17 Ganzzahlmultiplikationen). Der
Grund liegt darin, dass es bei Verwendung des RSA möglich ist, einen sehr
kurzen Exponent für den öffentlichen Schlüssel zu wählen, solange der
private Exponent nur hinreichend lang ist.

@@

Es gibt Faktorisierungsalgorithmen%
\footnote{John M. Pollard \index{Pollard, John M.} war an der
Entwicklung vieler verschiedener Faktorisierungsalgorithmen beteiligt; auch
beim Faktorisieren mit ECC war er einer der führenden Köpfe. Als Mitarbeiter
von British Telekom hat er leider nie viel selbst publiziert. Auf der RSA
Konferenz 1999 wurde er für seine \glqq outstanding contributions in
mathematics\grqq~ausgezeichnet.\\
Im Jahr 1987 stellte H.W. Lenstra\index{Lenstra 1987} einen häufig genutzten
Faktorisierungsalgorithmus vor, der auf Elliptischen Kurven basiert
(siehe \cite{Lenstra1987}).
  % Leerzeile vor schließender Footnote-Klammer zu #21 nötig, um einheitlich eine
  % Leerzeile am Ende der Fußnoten zu haben -- solange keine andere Lösung
  % für das Einrückproblem bei Fußnote #23 (TODO_LaTeX)

### 2018-05-20 script/en/chapters/introduction.tex
CT1: Updated CT book (d+e): cover page now with 2018 (before deleted cover with lower resolution for drafts); TeXlive 2018: commandline-make on Windows and WinEdt both worked well besides flaws documented in CT-Book-de.tex; removed babel warnings "empty language", added language in each bibliographies record; cleaned up the comments in both CT-Book-xx.txt and in reference.bib; fixed small typos, dead links, ...
CT1: Updated CT book (d+e): small typos

% $Id$
% ............................................................................
%      P R E F A C E  and  I N T R O D U C T I O N  (Playing together of book and programs)
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

%This edition of the CT book again updated some topics:
Compared to edition 11, this edition completely updated the TeX sources
of the document (e.g. one single bibtex file for all chapters and both languages),
and of course, the content of the book was amended, corrected, and
updated with many topics, for instance:
\vspace{-5pt}
\begin{itemize}
%  \item the definitions of the strength of security functions
%        (chapter \ref{cm_Section_Security_Definitions}),
  \item the largest prime numbers (chap. \ref{search_for_very_big_primes}),
%        new factorization records (chap. \ref{RSA-768}),
%  \item progress in cryptanalysis of AES
%        (chap. \ref{NeueAES-Analyse}) and
%  \item progress in cryptanalysis of hash algorithms
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor)
%        (chap. \ref{xxxxxxxxxcm_Brute-force-versus-Symmetr})\index{xxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory
        played a major role (see appendix \ref{s:appendix-movies}),
%       and where primes are used as hangers
%        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}),
  \item the overviews of all functions in
        \hyperlink{appendix-template-overview-CT2}{CrypTool~2 (CT2)},
        \hyperlink{appendix-function-overview-JCT}{JCrypTool (JCT)}, and
        \hyperlink{appendix-function-overview-CTO}{CrypTool-Online (CTO)}
        (see appendix),
  \item further SageMath scripts for cryptography, and the appendix
        \ref{s:appendix-using-sage} about using the computer algebra system
        SageMath,
        % Nguyen and Massierer (see also chap.~\ref{ec:Sage_Massierer}).
  \item the section about the Goldbach conjecture
        (see \ref{L-GoldbachConjecture}) and about twin primes
        (see \ref{L-TwinCousinPrimes}),
  \item the section about shared primes in RSA modules used in reality
        (see \ref{nt_Shared-Primes}),
%  \item the section about RSA fixed points
%        (see \ref{l:NumberTheory_Sage_Number-of-RSA-FixedPoints}), and
%  \item homomorphic encryption
%        (see chapter \ref{Chapter_HomomorphicCiphers}).
  \item the ``\nameref{Chapter_BitCiphers}'' is completely new
        (see chapter \ref{Chapter_BitCiphers}),
  \item the ``\nameref{Chapter_Dlog-FactoringDead}'' is completely new too
        (see chapter \ref{Chapter_Dlog-FactoringDead}). It's a phantastic
	in-depth summary about the limits of the according current
	cryptanalytical methods.
\end{itemize}

@@

\par \vskip + 35pt
\noindent Bernhard Esslinger
\par \vskip + 10pt
\noindent Heilbronn/Siegen (Germany), August 2016 + August 2017 + April 2018

\par \vskip + 90pt
\noindent PS:\\
We'd be glad if new authors would show up to improve existing chapters or to add further chapters,
e.g. about
\begin{compactitem}
   \item the Riemann Zeta function,
   \item hash functions and password attacks,
   \item lattice-based cryptography,
   \item random numbers,
   \item format-preserving encryption,
   \item privacy-preserving cryptography,
   \item the design and attack of crypto protocols (like SSL).
\end{compactitem}

@@

MTC3 is the abbreviation for MysteryTwister C3, an international cryptography
contest (\url{http://www.mysterytwisterc3.org}), which is also based on the
CT project.
Here you can find cryptographic riddles in four categories, a high-score list
and a moderated forum. As of 2016-06-16 more than 7000 users participate,
and more than 200 challenges are offered (162 of them are solved by
at least one participant).
\par \vskip + 25pt  % vskip: more than +15 to start SageMat the next page

>>>>

% $Id$
% ............................................................................
%      P R E F A C E  and  I N T R O D U C T I O N  (Playing together of book and programs)
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% WinEdt spell-check done 2018-05-21
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

%This edition of the CT book again updated some topics:
Compared to edition 11, this edition completely updated the TeX sources
of the document (e.g. one single bibtex file for all chapters and both languages),
and of course, the content of the book was amended, corrected, and
updated with many topics, for instance:
\vspace{-5pt}
\begin{itemize}
%  \item the definitions of the strength of security functions
%        (chapter \ref{cm_Section_Security_Definitions}),
  \item the largest prime numbers (chap. \ref{search_for_very_big_primes}),
%        new factorization records (chap. \ref{RSA-768}),
%  \item progress in cryptanalysis of AES
%        (chap. \ref{NeueAES-Analyse}) and
%  \item progress in cryptanalysis of hash algorithms
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor)
%        (chap. \ref{xxxxxxxxxcm_Brute-force-versus-Symmetr})\index{xxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory
        played a major role (see appendix \ref{s:appendix-movies}),
%       and where primes are used as hangers
%        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}),
  \item the overviews of all functions in
        \hyperlink{appendix-template-overview-CT2}{CrypTool~2 (CT2)},
        \hyperlink{appendix-function-overview-JCT}{JCrypTool (JCT)}, and
        \hyperlink{appendix-function-overview-CTO}{CrypTool-Online (CTO)}
        (see appendix),
  \item further SageMath scripts for cryptography, and the appendix
        \ref{s:appendix-using-sage} about using the computer algebra system
        SageMath,
        % Nguyen and Massierer (see also chap.~\ref{ec:Sage_Massierer}).
  \item the section about the Goldbach conjecture
        (see \ref{L-GoldbachConjecture}) and about twin primes
        (see \ref{L-TwinCousinPrimes}),
  \item the section about shared primes in RSA modules used in reality
        (see \ref{nt_Shared-Primes}),
%  \item the section about RSA fixed points
%        (see \ref{l:NumberTheory_Sage_Number-of-RSA-FixedPoints}), and
%  \item homomorphic encryption
%        (see chapter \ref{Chapter_HomomorphicCiphers}).
  \item the ``\nameref{Chapter_BitCiphers}'' is completely new
        (see chapter \ref{Chapter_BitCiphers}),
  \item the ``\nameref{Chapter_Dlog-FactoringDead}'' is completely new too
        (see chapter \ref{Chapter_Dlog-FactoringDead}). It's a fantastic
	in-depth summary about the limits of the according current
	cryptanalytic methods.
\end{itemize}

@@

\par \vskip + 35pt
\noindent Bernhard Esslinger
\par \vskip + 10pt
\noindent Heilbronn/Siegen (Germany), August 2016 + August 2017 + May 2018

\par \vskip + 90pt
\noindent PS:\\
We'd be glad if new authors would show up to improve existing chapters or to add further chapters,
e.g. about
\begin{compactitem}
   \item the Riemann Zeta function,
   \item hash functions and password attacks,
   \item lattice-based cryptography,
   \item random numbers,
   \item format-preserving encryption, privacy-preserving cryptography,
   \item discussion of the security effect of different block modes,
   \item the design and attack of crypto protocols (like SSL).
\end{compactitem}

@@

MTC3 is the abbreviation for MysteryTwister C3, an international cryptography
contest (\url{http://www.mysterytwisterc3.org}), which is also based on the
CT project.
Here you can find cryptographic riddles in four categories, a high-score list
and a moderated forum. As of 2016-06-16 more than 7000 users participate,
and more than 200 challenges are offered (162 of them are solved by
at least one participant).
\par \vskip + 25pt  % vskip: more than +15 to start SageMath the next page

### 2018-05-20 script/en/chapters/moderncryptography.tex
CT1: Updated CT book (d+e): cover page now with 2018 (before deleted cover with lower resolution for drafts); TeXlive 2018: commandline-make on Windows and WinEdt both worked well besides flaws documented in CT-Book-de.tex; removed babel warnings "empty language", added language in each bibliographies record; cleaned up the comments in both CT-Book-xx.txt and in reference.bib; fixed small typos, dead links, ...
CT1: Updated CT book (d+e): small typos

% $Id$
% ..........................................................................
%            The Mathematical Ideas behind Modern Cryptography
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

A number of algorithms have been suggested for public key procedures\index{cryptography!public key}. In many 
cases - although they at first appeared promising - it was discovered that they 
could be solved polynomially\index{polynomial}. The most famous failed applicant is the knapsack 
with trapdoor, suggested by Ralph Merkle \cite{Merkle1978}.

@@

%--------------------------------------------------------------------
\hypertarget{RSAVerfahren}{}
\subsection[The RSA procedure]{The RSA procedure\footnotemark$^,$\footnotemark}
\addtocounter {footnote} {-1} % Each footnotemark increase the footnotecounter by 1.
                              % So if there are two, we have to lower it bevor the first usage.
                              % And to raise it again before the 2nd usage.
\footnotetext{
  Please compare chapters \ref{rsabeweis}, ff.
}
%%_BEFORE_MOVING_OUT_INDEX: %%## }$^,$\footnote{%

@@

The current record\index{logarithm problem!record} (as of April 2007) for calculating discrete logarithms was
established in February 2007 by the group Kleinjung, Franke and Bahr at
University of Bonn.\footnote{%
   \url{http://www.nabble.com/Discrete-logarithms-in-GF(p)-----160-digits-t3175622.html}}
Kleinjung calculated the discrete logarithm modulo a 160 digit prime number $p$
and generator $g$:
$$
\begin{array}{r@{\:}c@{\:}l}
p & = & \lfloor 10^{159}\pi\rfloor + 119849\\
  & = & 314159265358979323846264338327950288419716939937510582097494\\
  &   & 459230781640628620899862803482534211706798214808651328230664\\
  &   & 7093844609550582231725359408128481237299\\
g & = & 2
\end{array}
$$
The discrete logarithms $k$ of the following integer $y$ was
determined:\footnote{The integer $y$ was chosen as the first 159 digits of the
Euler number $e$.}
$$
\begin{array}{r@{\:}c@{\:}l}
y & = & \lfloor 10^{159}e\rfloor\\
  & = & 271828182845904523536028747135266249775724709369995957496696\\
  &   & 762772407663035354759457138217852516642742746639193200305992\\
  &   & 1817413596629043572900334295260595630738\\
k & = & \log_g(y)\mod p\\
  & = & 829897164650348970518646802640757844024961469323126472198531\\
&   & 845186895984026448342666252850466126881437617381653942624307\\
&   & 537679319636711561053526082423513665596
\end{array}
$$
The search was performed with GNFS method (general number field sieve,
Index-Calculus) \index{general number field sieve (GNFS)} and took about
17 CPU years on 3.2 GHz Xeon machines.

>>>>

% $Id$
% ..........................................................................
%            The Mathematical Ideas behind Modern Cryptography
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% WinEdt spell-check done 2018-05-21
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

A number of algorithms have been suggested for public key procedures\index{cryptography!public key}. In many
cases -- although they at first appeared promising -- it was discovered that they
could be solved polynomially\index{polynomial}. The most famous failed applicant is the knapsack
with trapdoor, suggested by Ralph Merkle \cite{Merkle1978}.

@@

%--------------------------------------------------------------------
\hypertarget{RSAVerfahren}{}
\subsection[The RSA procedure]{The RSA procedure\footnotemark$^,$\footnotemark}
\addtocounter {footnote} {-1} % Each footnotemark increase the footnotecounter by 1.
                              % So if there are two, we have to lower it before the first usage.
                              % And to raise it again before the 2nd usage.
\footnotetext{
  Please compare chapters \ref{rsabeweis}, ff.
}
%%_BEFORE_MOVING_OUT_INDEX: %%## }$^,$\footnote{%

@@

The current record\index{logarithm problem!record} (as of April 2007) for calculating discrete logarithms was
established in February 2007 by the group Kleinjung, Franke, and Bahr at
University of Bonn.\footnote{%
   \url{http://www.nabble.com/Discrete-logarithms-in-GF(p)-----160-digits-t3175622.html}}
Kleinjung calculated the discrete logarithm modulo a 160 digit prime number $p$
and generator $g$:
$$
\begin{array}{r@{\:}c@{\:}l}
p & = & \lfloor 10^{159}\pi\rfloor + 119849\\
  & = & 314159265358979323846264338327950288419716939937510582097494\\
  &   & 459230781640628620899862803482534211706798214808651328230664\\
  &   & 7093844609550582231725359408128481237299\\
g & = & 2
\end{array}
$$
The discrete logarithms $k$ of the following integer $y$ was
determined:\footnote{The integer $y$ was chosen as the first 159 digits of the
Euler number $e$.}
$$
\begin{array}{r@{\:}c@{\:}l}
y & = & \lfloor 10^{159}e\rfloor\\
  & = & 271828182845904523536028747135266249775724709369995957496696\\
  &   & 762772407663035354759457138217852516642742746639193200305992\\
  &   & 1817413596629043572900334295260595630738\\
k & = & \log_g(y)\mod p\\
  & = & 829897164650348970518646802640757844024961469323126472198531\\
&   & 845186895984026448342666252850466126881437617381653942624307\\
&   & 537679319636711561053526082423513665596
\end{array}
$$
The search was performed with GNFS method (general number field sieve,
Index-Calculus) \index{general number field sieve (GNFS)} and took about
17 CPU years on 3.2 GHz Xeon machines.

### 2018-05-20 script/en/chapters/numbertheory.tex
CT1: Updated CT book (d+e): small typos

% $Id$
% ..........................................................................
%              E l e m e n t a r y  N u m b e r  T h e o r y
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

\noindent The following three examples\footnote{%
  See chapter~\ref{nt:AppArith1} ``\nameref{nt:AppArith1}''
  for the source code to compute the tables using SageMath.}
illustrate the properties of multiplicative inverses (here only the lines
for the factors $5$ und $6$ are listed; not the complete multiplication table).

@@

\noindent a) If the multiplicative order exists for $a$, (indendently whether
$a$ is prim) it is: $ord_{n}(a) = l$.
\indent Samples: The maximum length $l_{max}$\footnote{%
            We don't know of a formular telling for which a the length has a maximum.}
is achieved e.g. for:\\
\indent  - $a=3$ with $l_{max} = ord_{14}(a) = 6$ in table~\ref{expmod14}, or\\
\indent  - $a=10$ with $l_{max} = ord_{22}(a) = 10$ in table~\ref{expmod22}.

@@

Experiences about the ellipsed time of factorization\index{factorization}
with the open source software Pari-GP, SageMath, CrypTool~1 and CrypTool~2)
can be found in ``Zeitexperimente zur Faktorisierung'' (time experiments
about factorization) (see \cite{Schulz2010}).

@@

In addition, the operations multiplication, divsion and gcd show big differences in runtime when the used operands are very different in size.

@@

\noindent As the SageMath function discrete\_log expects as arguments
only elements of a ring (integers between 0 and an upper limit),
we could either enforce this type by entering the numbers directly with the
according modulo operator: \texttt{discrete\_log(mod(49, 347), mod(11, 347))}.\\
A much better alternative is to use the variables like in the formula above
and let SageMath know that they are elements of a ring.
After this ``burdon'' at the beginning for the initialization, you can write
the formulars like you are used to: \texttt{discrete\_log(A, g)}.\footnote{%
Such number theoretic tasks can also be solved using other
tools like PariGP\index{Pari-GP}, LiDIA\index{LiDIA}, BC\index{BC}, or
Mathematica\index{Mathematica}
(see the list of web sites in the appendix at the end of this chapter).
Here is the according syntax to get the discrete log for Alice:
	\begin{compactitem}
	\item Pari-GP: \texttt{znlog(Mod(49,347),Mod(11,347))}
	\item LiDIA:   \texttt{dl(11,49,347)}
	\item Mathematica: {\tt MultiplicativeOrder[11, 347, 49]}\\
		The general ``Solve'' function provides the ``{em tdep message}'':
		The equations appear to involve the variables to be solved
		for in an essentially non-algebraic way.
	\end{compactitem}
All function calls deliver the result $67$.
}${}^,$\footnote{%
Why have the functions delivered the value $67$ for the discrete
logarithm\index{discrete logarithm} of Alice rather than $240$ which Alice
selected as exponent $a$?\\
The discrete logarithm is
the smallest natural exponent that solves the equation 
$11^x \equiv 49 \pmod{347}$. Both $x = 67$ and $x = 240$ (the number
selected in the example) satisfy the equation and can therefore be
used to calculate the session key: 
$285^{240} \equiv 285^{67} \equiv 268 \pmod{347}$.
If Alice and Bob
had selected a primitive root\index{primitive root} modulo $p$ as base $g$, then for every
remainder from the set $\{1, 2, \dots, p-1 \}$ there is exactly one
exponent from the set $\{0, 1, \dots, p-2 \}$.\\
\indent As an aside, there are $172$ different primitive roots modulo $347$,
$32$ of which are prime (not necessary). Since the number $11$
selected for $g$ in the example is not a primitive root\index{primitive root} of $347$, the
remainders do not take all values from the set $\{1, 2, \dots, 346 \}$. 
Thus, for a particular remainder there may be more than one exponent
or even no exponent at all in the set $\{0, 1, \dots, 345 \}$ that
satisfies the equation.\\
With the relevant SageMath\index{SageMath} commands you find:\\
\texttt{is\_prime(347)=True}, \texttt{euler\_phi(347)=346}, \texttt{gcd(11,347)=1}, and 
\texttt{multiplicative\_order(mod(11, 347))=173}.

@@

There are three tasks with completely different degrees of difficulty here.
In each case we know the ciphertext 
and the public key $(e,n)$:
\begin{itemize}
\item[{\bf (a)}] Known plaintext: \index{attack!known plaintext} find the secret key $d$ using the additionally known original message.
\item[{\bf (b)}] Ciphertext-only: \index{attack!ciphertext-only} find $d$ and the plaintext.
\item[{\bf (c)}] Calculate the RSA modulus, in other words factorization (with no knowledge of the message). \index{factorization!factorization problem}
\end{itemize}

@@

The folowing presentation (last update Nov. 2010) shows the basics of the
RSA cipher in an interactive way.

@@

# AA_Start -- Testcases for enum_PrimitiveRoots_of_an_Integer(M)
print "AA_Start -- Testcases for enum_PrimitiveRoots_of_an_Integer(M)"
M=10; print "1-----------Testcase: M = %s" % M
LL = enum_PrimitiveRoots_of_an_Integer(M)
if LL==None:
    print M
else:
    print LL
M=8; print "2-----------Testcase: M = %s" % M
# M=8 hat keine primitive root mod m. Checke, ob per try - except abgefangen.
LL = enum_PrimitiveRoots_of_an_Integer(M)
if LL==None:
    print M
else:
    print LL
M=17; print "3-----------Testcase: M = %s" % M
LL = enum_PrimitiveRoots_of_an_Integer(M)
if LL==None:
    print M
else:
    print LL
# AA_End -- Testcases

@@

Also encryption methods can have fixed -- cleartext messages where the
according ciphertext matches the original. In mathematics, variables
mapped by the algorithm (function) onto themselves are called fixed points.
In cryptography the according messages are called ``unconcealed messages''.

@@

With the RSA procedure: $n=pq$ is the product of two different prime numbers,
and there exists $e$ where $gcd(e,(p-1)(q-1))=1$. The encription is then
$c = m^e \mod n$. 
A fixed point in the RSA procedure is a message $m$, where: 
$m = m^e \mod n$. 
The result of the encryption is the given message.

When the size of $n$ is sufficiently big, the probability of the occurance of
fixed points in RSA is very small -- as illustrated in Figure
\ref{fig:NumberFixpointsGrowingN}: In average, we found not more than 40 fixed
points.

Students often presume the occurence of fixed points high, because they counter
a ``relatively'' large number of examples when experimenting with \textbf{small}
prime numbers, as m = 0, 1 and n-1 are also always fixed points.

@@

~\\  
\noindent There is not just the one single $e$ (see above), where all
$ m \in \{1,...,n-1\}$ have the fixed point property
$ m^e = m \bmod n$.\footnote{%
  Sometimes these $e$, which make any message to a fixed point, are called
  ``weak keys'' $(e,n)$ of the RSA algorithm\index{key!weak}.
  This notation is different to the ``weak keys'' $k$ in DES\index{DES}, where
  \textbf{every} message $m$ relates to itself if the \textbf{en}cryption is
  done twice.
  To my knowledge, for larger $n$ the RSA procedure does not have weaks in this
  meaning: $(m^e)^e = m$.\\
  In JCT\index{JCT} you can find weak DES keys in the default perspective
  via the menu item {\bf Visuals \textbackslash{} Inner States of the Data
  Encryption Standard (DES)}.
}

@@

   \item BC \index{BC}\\
         Interpreter with number-theory functions. Development stopped since 2006.
	 The existing BC versions don't work correctly under new Windows versions.\\
	 Keith Matthews regularily publishes updates for UNIX.\\
         \url{http://www.gnu.org/software/bc/bc.html}\\
	 \url{http://www.numbertheory.org/gnubc/gnubc.html}

>>>>

% $Id$
% ..........................................................................
%              E l e m e n t a r y  N u m b e r  T h e o r y
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% WinEdt spell-check done 2018-05-21
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

@@

\noindent The following three examples\footnote{%
  See chapter~\ref{nt:AppArith1} ``\nameref{nt:AppArith1}''
  for the source code to compute the tables using SageMath.}
illustrate the properties of multiplicative inverses (here only the lines
for the factors $5$ and $6$ are listed; not the complete multiplication table).

@@

\noindent a) If the multiplicative order exists for $a$ (independently whether
$a$ is prim), it is: $ord_{n}(a) = l$.
\indent Samples: The maximum length $l_{max}$\footnote{%
            We don't know of a formula telling for which a the length has a maximum.}
is achieved e.g. for:\\
\indent  - $a=3$ with $l_{max} = ord_{14}(a) = 6$ in table~\ref{expmod14}, or\\
\indent  - $a=10$ with $l_{max} = ord_{22}(a) = 10$ in table~\ref{expmod22}.

@@

Experiences about the elapsed time of factorization\index{factorization}
with the open source software Pari-GP, SageMath, CrypTool~1 and CrypTool~2)
can be found in ``Zeitexperimente zur Faktorisierung'' (time experiments
about factorization) (see \cite{Schulz2010}).

@@

Furthermore, the operations multiplication, division and gcd show big differences in runtime when the used operands are very different in size.

@@

\noindent As the SageMath function discrete\_log expects as arguments
only elements of a ring (integers between 0 and an upper limit),
we could either enforce this type by entering the numbers directly with the
according modulo operator: \texttt{discrete\_log(mod(49, 347), mod(11, 347))}.\\
A much better alternative is to use the variables like in the formula above
and let SageMath know that they are elements of a ring.
After this ``burdon'' at the beginning for the initialization, you can write
the formulas like you are used to: \texttt{discrete\_log(A, g)}.\footnote{%
Such number theoretic tasks can also be solved using other
tools like PariGP\index{Pari-GP}, LiDIA\index{LiDIA}, BC\index{BC}, or
Mathematica\index{Mathematica}
(see the list of web sites in the appendix at the end of this chapter).
Here is the according syntax to get the discrete log for Alice:
	\begin{compactitem}
	\item Pari-GP: \texttt{znlog(Mod(49,347),Mod(11,347))}
	\item LiDIA:   \texttt{dl(11,49,347)}
	\item Mathematica: {\tt MultiplicativeOrder[11, 347, 49]}\\
		The general ``Solve'' function provides the ``{em tdep message}'':
		The equations appear to involve the variables to be solved
		for in an essentially non-algebraic way.
	\end{compactitem}
All function calls deliver the result $67$.
}${}^,$\footnote{%
Why have the functions delivered the value $67$ for the discrete
logarithm\index{discrete logarithm} of Alice rather than $240$ which Alice
selected as exponent $a$?\\
The discrete logarithm is
the smallest natural exponent that solves the equation
$11^x \equiv 49 \pmod{347}$. Both $x = 67$ and $x = 240$ (the number
selected in the example) satisfy the equation and can therefore be
used to calculate the session key:
$285^{240} \equiv 285^{67} \equiv 268 \pmod{347}$.
If Alice and Bob
had selected a primitive root\index{primitive root} modulo $p$ as base $g$, then for every
remainder from the set $\{1, 2, \dots, p-1 \}$ there is exactly one
exponent from the set $\{0, 1, \dots, p-2 \}$.\\
\indent As an aside, there are $172$ different primitive roots modulo $347$,
$32$ of which are prime (not necessary). Since the number $11$
selected for $g$ in the example is not a primitive root\index{primitive root} of $347$, the
remainders do not take all values from the set $\{1, 2, \dots, 346 \}$.
Thus, for a particular remainder there may be more than one exponent
or even no exponent at all in the set $\{0, 1, \dots, 345 \}$ that
satisfies the equation.\\
With the relevant SageMath\index{SageMath} commands you find:\\
\texttt{is\_prime(347)=True}, \texttt{euler\_phi(347)=346}, \texttt{gcd(11,347)=1}, and
\texttt{multiplicative\_order(mod(11, 347))=173}.

@@

There are three tasks with completely different degrees of difficulty here.
In each case we know the ciphertext
and the public key $(e,n)$:
\begin{itemize}
\item[{\bf (a)}] Known plaintext: \index{attack!known plaintext} find the secret key $d$ using the additionally known original message.
\item[{\bf (b)}] Ciphertext-only: \index{attack!ciphertext-only} find $d$ and the plaintext.
\item[{\bf (c)}] Calculate the RSA modulus, in other words factorization\footnote{%
  Here, splitting n into prime integers produces: $n = 63978486879527143858831415041 = 145295143558111 * 440334654777631$.
} (with no knowledge of the message). \index{factorization!factorization problem}
\end{itemize}

@@

The following presentation (last update Nov. 2010) shows the basics of the
RSA cipher in an interactive way.

@@

# AA_Start -- Testcases for enum_PrimitiveRoots_of_an_Integer(M)
print "AA_Start -- Testcases for enum_PrimitiveRoots_of_an_Integer(M)"
M=10; print "1-----------Testcase: M = %s" % M
LL = enum_PrimitiveRoots_of_an_Integer(M)
if LL==None:
    print M
else:
    print LL
M=8; print "2-----------Testcase: M = %s" % M
# M=8 has no primitive root mod m. You could check via try - except.
LL = enum_PrimitiveRoots_of_an_Integer(M)
if LL==None:
    print M
else:
    print LL
M=17; print "3-----------Testcase: M = %s" % M
LL = enum_PrimitiveRoots_of_an_Integer(M)
if LL==None:
    print M
else:
    print LL
# AA_End -- Testcases

@@

Also encryption methods can have fixed plaintext messages where the
according ciphertext matches the original. In mathematics, variables
mapped by the algorithm (function) onto themselves are called fixed points.
In cryptography the according messages are called ``unconcealed messages''.

@@

With the RSA procedure: $n=pq$ is the product of two different prime numbers,
and there exists $e$ where $gcd(e,(p-1)(q-1))=1$. The encryption is then
$c = m^e \mod n$.
A fixed point in the RSA procedure is a message $m$, where:
$m = m^e \mod n$.
The result of the encryption is the given message.

When the size of $n$ is sufficiently big, the probability of the occurrence of
fixed points in RSA is very small -- as illustrated in Figure
\ref{fig:NumberFixpointsGrowingN}: In average, we found not more than 40 fixed
points.

Students often presume the occurrence of fixed points high, because they counter
a ``relatively'' large number of examples when experimenting with \textbf{small}
prime numbers, as m = 0, 1 and n-1 are also always fixed points.

@@

~\\
\noindent There is not just the one single $e$ (see above), where all
$ m \in \{1,...,n-1\}$ have the fixed point property
$ m^e = m \bmod n$.\footnote{%
  Sometimes these $e$, which make any message to a fixed point, are called
  ``weak keys'' $(e,n)$ of the RSA algorithm\index{key!weak}.
  This notation is different to the ``weak keys'' $k$ in DES\index{DES}, where
  \textbf{every} message $m$ relates to itself if the \textbf{en}cryption is
  done twice.
  To my knowledge, for larger $n$ the RSA procedure does not have weaknesses in this
  meaning: $(m^e)^e = m$.\\
  In JCT\index{JCT} you can find weak DES keys in the default perspective
  via the menu item {\bf Visuals \textbackslash{} Inner States of the Data
  Encryption Standard (DES)}.
}

@@

   \item BC \index{BC}\\
         Interpreter with number-theory functions. Development stopped since 2006.
	 The existing BC versions don't work correctly under new Windows versions.\\
	 Keith Matthews regularly publishes updates for UNIX.\\
         \url{http://www.gnu.org/software/bc/bc.html}\\
	 \url{http://www.numbertheory.org/gnubc/gnubc.html}
